Compiling refinement types to dependent types

### 2020-07-06 tex/paper.tex
Something weird with spacing there
We'll need equality as well
Using \overline for the core language is a really bad idea
We'll probably use it later to denote a sequence of terms/types in the
case-split rule.
Show the syntax for the contexts as well
Split surface language typing and subtyping
Something funky with spaces again
We actually have two key theorems

\newNTclass{nonterm}
\newnonterm e \varepsilon
\newnonterm{es}{\varepsilon}
\newnonterm{ec}{\overline\varepsilon}
\newnonterm r \rho
\newnonterm{ts}{\tau}
\newnonterm s s
\newnonterm B B
\newnonterm G \Gamma
\newnonterm{GC}{\overline\Gamma}
\newnonterm{binop}{\odot}

@@

\begin{figure}[ht]
  \footnotesize
  \begin{subfigure}{.6\textwidth}
	\nonterms{es}
	\caption{Term level}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
	\nonterms{r,B,ts}
	\caption{Type level}
  \end{subfigure}
  \caption{Surface language syntax}
  \label{fig:surface_syntax}
\end{figure}

\begin{figure}[ht]
  \footnotesize
  \drules[TWF]{$\nonterm G \vdash \nonterm{ts} \text{ ok}$}{type well-formedness}{TrueRef,Base,Conj,Arr}
  \drules[T]{$\nonterm G \vdash \nonterm{es} : \nonterm{ts}$}{term typing}{BoolTop,BoolBot,Var,Abs,App,If,Sub}
  \drules[ST]{$\nonterm G \vdash \nonterm{ts}_1 <: \nonterm{ts}_2$}{subtyping}{Base,Arr}
  \caption{Surface language typing rules}
  \label{fig:surface_typing}
\end{figure}

The translation function $[|\ |]$ maps the typing context $\nonterm G$ to some logic:
\begin{equation}
\begin{aligned}
  & [| \emptyset |]                                         && \triangleq && \top																						  \\
  & [| \nonterm G, x : \{ v : \nonterm B | \nonterm r \} |] && \triangleq && [| \nonterm G |] \land [| \{x/v\} \nonterm r |]  \\
  & [| \nonterm r_1 \land \nonterm r_2 |]                   && \triangleq && [| \nonterm r_1 |] \land [| \nonterm r_2 |]      \\
  & [| \nonterm{es} |]                                      && \triangleq && \text{oracle}(\nonterm{es})                      \\
  & [| \nonterm{es}_1 = \nonterm{es}_2 |]                   && \triangleq && \text{oracle}(\nonterm{es}_1 = \nonterm{es}_2)   \\
\end{aligned}
\end{equation}

@@

\begin{itemize}
  \item Non-dependent function type:
  \[
    \tau_1 \rightarrow \tau_2 \triangleq \Pi \_ : \tau_1. \tau_2.
  \]
  \item Dependent pair type, or $\Sigma$-type, with the first component being a value $x$ of type $\tau$ and the second component being of type $P x$:
  \[
    \{ x : \tau\ **\ P x \} \triangleq \Pi \alpha : \star. (\Pi x : \tau. P x \rightarrow \alpha) \rightarrow \alpha.
  \]
  \item Accessing the first and second component of the dependent pair:
  \[
    \text{fst} \triangleq TODO,
  \]
  \[
    \text{snd} \triangleq TODO.
  \]
  \item Non-dependent pair type:
  \[
    (\tau_1, \tau_2) \triangleq \{ x : \tau_1\ **\ (\lambda \_ : \tau_1. \tau_2) x \}.
  \]
\end{itemize}

@@

We next formulate our key theorem:

\begin{theorem}
  For any well-typed surface language term $\nonterm{es}$,
  the result of the translation $\mu_\epsilon(\nonterm{es})$ is well-typed in the core language:
  \[
    \nonterm G \vdash \nonterm{es} : \nonterm{ts} \implies \mu_\Gamma(\nonterm G) \vdash \mu_\epsilon(\nonterm{es}) : \mu_\tau(\nonterm{ts})
  \]
\end{theorem}

>>>>

\newNTclass{nonterm}
\newnonterm e \varepsilon
\newnonterm{es}{\varepsilon}
\newnonterm{ec}{\hat\varepsilon}
\newnonterm r \rho
\newnonterm{ts}{\tau}
\newnonterm s s
\newnonterm B B
\newnonterm G \Gamma
\newnonterm{GC}{\hat\Gamma}
\newnonterm{binop}{\odot}

@@

\begin{figure}[ht]
  \footnotesize
  \begin{subfigure}{.6\textwidth}
    \nonterms{es}
    \caption{Term level}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
    \nonterms{G,r,B,ts}
    \caption{Type level}
  \end{subfigure}
  \caption{Surface language syntax}
  \label{fig:surface_syntax}
\end{figure}

\begin{figure}[ht]
  \footnotesize
  \drules[TWF]{$\nonterm G \vdash \nonterm{ts} \text{ ok}$}{type well-formedness}{TrueRef,Base,Conj,Arr}
  \drules[T]{$\nonterm G \vdash \nonterm{es} : \nonterm{ts}$}{term typing}{BoolTop,BoolBot,Var,Abs,App,If,Sub}
  \caption{Surface language typing}
  \label{fig:surface_typing}
\end{figure}

\begin{figure}[ht]
  \footnotesize
  \drules[ST]{$\nonterm G \vdash \nonterm{ts}_1 <: \nonterm{ts}_2$}{subtyping}{Base,Arr}
  \caption{Surface language subtyping}
  \label{fig:surface_subtyping}
\end{figure}
The translation function $[|\ |]$ maps the typing context $\nonterm G$ to some logic:
\begin{equation}
\begin{aligned}
  & [| \emptyset |]                                         && \triangleq && \top																						  \\
  & [| \nonterm G, x : \{ v : \nonterm B | \nonterm r \} |] && \triangleq && [| \nonterm G |] \land [| \{x/v\} \nonterm r |]  \\
  & [| \nonterm r_1 \land \nonterm r_2 |]                   && \triangleq && [| \nonterm r_1 |] \land [| \nonterm r_2 |]      \\
  & [| \nonterm{es} |]                                      && \triangleq && \text{oracle}(\nonterm{es})                      \\
  & [| \nonterm{es}_1 = \nonterm{es}_2 |]                   && \triangleq && \text{oracle}(\nonterm{es}_1 = \nonterm{es}_2)   \\
\end{aligned}
\end{equation}

@@

\begin{itemize}
  \item Non-dependent function type:
    \[
      \tau_1 \rightarrow \tau_2 \triangleq \Pi \_ : \tau_1. \tau_2.
    \]
  \item Dependent pair type, or $\Sigma$-type, with the first component being a value $x$ of type $\tau$ and the second component being of type $P x$:
    \[
      \{ x : \tau\ **\ P x \} \triangleq \Pi \alpha : \star. (\Pi x : \tau. P x \rightarrow \alpha) \rightarrow \alpha.
    \]
  \item Accessing the first and second component of the dependent pair:
    \[
      \text{fst} \triangleq TODO,
    \]
    \[
      \text{snd} \triangleq TODO.
    \]
  \item Non-dependent pair type:
    \[
      (\tau_1, \tau_2) \triangleq \{ x : \tau_1\ **\ (\lambda \_ : \tau_1. \tau_2) x \}.
    \]
  \item Equality of expressions $\nonterm{ec}_1, \nonterm{ec}_2$ of type $S$:
    \[
      \nonterm{ec}_1 \equiv \nonterm{ec}_2 \triangleq \Pi P : S \rightarrow \star. (P x \rightarrow P y, P y \rightarrow P x).
    \]
    Careful reader will recognize this as the Leibniz equality \cite{FindSomethingForThis}.
\end{itemize}

@@

The correctness of the translation hinges on two key theorems.
The first one claims that a well-typed program in surface language remains well-typed after the translation to the core language.
The second one claims that a well-typed surface language program evaluates to ``the same'' value as the translated one.

\begin{theorem}[Well-typedness preservation]
  For any well-typed surface language term $\nonterm{es}$,
  the result of the translation $\mu_\epsilon(\nonterm{es})$ is ``equivalently'' typeable in the core language:
  \[
    \nonterm G \vdash \nonterm{es} : \nonterm{ts} \implies \mu_\Gamma(\nonterm G) \vdash \mu_\epsilon(\nonterm{es}) : \mu_\tau(\nonterm{ts})
  \]
\end{theorem}
\begin{proof}
\end{proof}

\begin{theorem}[Evaluation equivalence]
  For any well-typed surface language term $\nonterm{es}$, if $\nonterm{es}$ evaluates to v, then $\mu_\varepsilon(\nonterm{es})$ evaluates to $\mu_\varepsilon(v)$:
  %TODO symbols and all that for v
\end{theorem}
\begin{proof}
\end{proof}
